---
title: "CytAssist FFPE multi-sample analysis"
description: |
author:
  - name: Mei Du
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_float: true
    code_folding: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(here)
library(SpatialExperiment)
library(Seurat)
library(ggplot2)
library(tidyverse)
library(patchwork)
```


# Notes

nnSVG feature selection + iSC.MEB downstream analysis:

* [nnSVG](https://www.biorxiv.org/content/10.1101/2022.05.16.492124v3)
  + "For datasets consisting of multiple biological samples (e.g. multiple 10x Genomics Visium capture areas, or multiple tissue samples from other platforms), we can run nnSVG once per sample and combine the results by averaging the ranks of the SVGs identified per sample."

* [iSC.MEB](https://academic.oup.com/bioinformaticsadvances/article/3/1/vbad019/7043976)
  + "First, we perform principal component analysis (PCA) on the combined log-transformed expression matrix to obtain the top principal components (PCs). Second, by taking the top PCs as input, we perform the integrative spatial clustering analysis using iSC-MEB that produces the soft cluster assignment as output. Finally, we can perform downstream analysis using the clustering result from the second step."

# Load data

```{r}
# pre-processed SpatialExperiment objects
spe_709 <- readRDS(here("analysis","output","RDS","spe_709_FFPE_CA_norm.RDS"))
spe_713 <- readRDS(here("analysis","output","RDS","spe_713_FFPE_CA_norm.RDS"))

# Combining samples with SpatialExperiment's cbind()
spe_709$sample_id <- "709"
spe_713$sample_id <- "713"
spe <- cbind(spe_709, spe_713)
table(colData(spe)$sample_id)

# HVGs
hvg_709 <- read.csv(here("analysis","output","HVGs","709_FFPE_CA_top_HVGs.csv"))
hvg_713 <- read.csv(here("analysis","output","HVGs","713_FFPE_CA_top_HVGs.csv"))
```


# nnSVG

Run with included R and SLURM scripts.

```{r,eval=FALSE}
library(nnSVG)

# Run nnSVG once per sample and store lists of top SVGs
sample_ids <- unique(colData(spe)$sample_id)

res_list <- as.list(rep(NA, length(sample_ids)))
names(res_list) <- sample_ids
colnames(rowData(spe)) <- c("gene_name","Chr")

for (s in seq_along(sample_ids)) {
  
  # select sample
  ix <- colData(spe)$sample_id == sample_ids[s]
  spe_sub <- spe[, ix]
  
  # run nnSVG filtering for mitochondrial genes and low-expressed genes
  spe_sub <- filter_genes(
    spe_sub, 
    filter_genes_ncounts = 3, 
    filter_genes_pcspots = 0.5, 
    filter_mito = TRUE
  )
  
  # remove any zeros introduced by filtering
  ix_zeros <- colSums(counts(spe_sub)) == 0
  if (sum(ix_zeros) > 0) {
    spe_sub <- spe_sub[, !ix_zeros]
  }
  
  dim(spe_sub)
  
  # re-calculate logcounts after filtering
  spe_sub <- computeLibraryFactors(spe_sub)
  spe_sub <- logNormCounts(spe_sub)
  
  
  # run nnSVG
  set.seed(123)
  spe_sub <- nnSVG(spe_sub)
  
  # store results for this sample
  res_list[[s]] <- rowData(spe_sub)
}

# saveRDS(res_list, here("analysis","output","RDS","nnSVG_FFPE_CA.RDS"))
```

Now that we have run nnSVG once per sample, we can combine the results across multiple samples by averaging the ranks of the SVGs. Moran’s I statistic is used to rank genes as SVGs, with the highest values (close to +1) representing the top-ranked SVGs.

```{r}
res_list <- readRDS(here("analysis","output","RDS","nnSVG_FFPE_CA.RDS"))

sample_ids <- unique(colData(spe)$sample_id)

# number of genes that passed filtering for each sample
sapply(res_list, nrow)

# add another filter - by significance
res_list_sig <- lapply(res_list, function(df) {
    return(df[df$padj < 0.05,])
})

# number of genes that passed significance
sapply(res_list_sig, nrow)

# match results from each sample and store in matching rows
res_ranks <- matrix(NA, nrow = nrow(spe), ncol = length(sample_ids))
rownames(res_ranks) <- rownames(spe)
colnames(res_ranks) <- sample_ids

for (s in seq_along(sample_ids)) {
  stopifnot(colnames(res_ranks)[s] == sample_ids[s])
  stopifnot(colnames(res_ranks)[s] == names(res_list)[s])
  
  rownames_s <- rownames(res_list[[s]])
  res_ranks[rownames_s, s] <- res_list[[s]][, "rank"]
}

# remove genes that were filtered out in all samples
ix_allna <- apply(res_ranks, 1, function(r) all(is.na(r)))
res_ranks <- res_ranks[!ix_allna, ]

dim(res_ranks)

# calculate average ranks
# note missing values due to filtering for samples
avg_ranks <- rowMeans(res_ranks, na.rm = TRUE)


# calculate number of samples where each gene is within top 100 ranked SVGs
# for that sample
n_withinTop100 <- apply(res_ranks, 1, function(r) sum(r <= 100, na.rm = TRUE))


# summary table
df_summary <- data.frame(
  gene_id = names(avg_ranks), 
  gene_name = rowData(spe)[names(avg_ranks), "symbol"], 
  #gene_type = rowData(spe)[names(avg_ranks), "gene_type"], 
  overall_rank = rank(avg_ranks), 
  average_rank = unname(avg_ranks), 
  n_withinTop100 = unname(n_withinTop100), 
  row.names = names(avg_ranks)
)

# sort by average rank
df_summary <- df_summary[order(df_summary$average_rank), ]
head(df_summary)

# remove genes that were filtered out in all samples
ix_allna <- apply(res_ranks, 1, function(r) all(is.na(r)))
res_ranks <- res_ranks[!ix_allna, ]

dim(res_ranks)

# top n genes
# (note: NAs in this example due to subsampling genes for faster runtime)
top100genes <- df_summary$gene_name[1:100]

# summary table of "replicated" SVGs (i.e. genes that are highly ranked in at
# least x samples)
df_summaryReplicated <- df_summary[df_summary$n_withinTop100 >= 2, ]

# re-calculate rank within this set
df_summaryReplicated$overall_rank <- rank(df_summaryReplicated$average_rank)

dim(df_summaryReplicated)

# top "replicated" SVGs
topSVGsReplicated <- df_summaryReplicated$gene_name

head(df_summary, 5)

# saveRDS(df_summary, here("analysis","output","RDS","FFPE_CA_SVGs_.RDS"))
```

## Gene annotation

```{r}
svg <- readRDS(here("analysis","output","RDS","FFPE_CA_SVGs.RDS"))

library(biomaRt)
mart <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")
svg_anno <- biomaRt::select(mart, keys=rownames(svg), keytype="external_gene_name",columns=c("external_gene_name","entrezgene_id"))
colnames(svg_anno)[1] <- "gene_name"
svg <- left_join(svg, svg_anno) %>% .[!is.na(.$entrezgene_id),]

# duplicated
svg[duplicated(svg[,"gene_name"]) | duplicated(svg[,"gene_name"], fromLast = TRUE),]

library(clusterProfiler)
library(org.Mm.eg.db)
go <- enrichGO(gene = as.character(svg$entrezgene_id)[1:1000], 
              OrgDb = org.Mm.eg.db,
              ont = "BP",
              pvalueCutoff = 0.05,
              universe = as.character(svg$entrezgene_id),
              readable =TRUE)
# pdf(here("analysis","output","top1000_SVGs_GO_heatmap.pdf"),height=5)
dotplot(go) + scale_colour_continuous(labels=scales::scientific_format(digits=1),low="royalblue3",high="orange")
# dev.off()
```

## Spatial SVG expression

```{r}
# filter out haemoglobin/immunoglobin genes from top SVGs
df_summary <- df_summary[-c(1:3),]
# ix_gene <- c("Car2","Tmcc2") # use to plot top 2 HVGs in figure 3a

# plot top genes
for (i in 1:15) {
  ix_gene <- which(rowData(spe)$symbol %in% df_summary[i, "gene_name"])

  df <- as.data.frame(cbind(
  colData(spe), 
  spatialCoords(spe), 
  gene = counts(spe)[ix_gene, ]))

  plot <- ggplot(df, aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = gene)) + 
  facet_wrap(~ sample_id, nrow = 1, scales = "free") + 
  geom_point(size = 0.6) + 
  scale_color_gradient(low = "gray80", high = "red", trans = "sqrt", 
                       name = "counts", breaks = range(df$gene)) + 
  scale_y_reverse() + 
  ggtitle(paste0(i,". ",rowData(spe)$symbol[ix_gene])) + 
  theme_bw() + 
  theme(panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        strip.background = element_rect(fill = "#DCEFFE", color = "#DCEFFE"),
                          strip.text = element_text(color = "#074D6C", face = "bold", size = 15))

  assign(paste0("p",i), plot)
}

# pdf(here("analysis","output","FFPE_CA_top2_SVGs.pdf"),height=2.3, width=7)
p1+p2
# dev.off()
```

```{r, fig.cap="The top 9 spatially variable genes across all samples.",layout="l-body-outset",fig.width=10, fig.height=10}
p1+p2+p3+p4+p5+p6+p7+p8+p9
```


# iSC.MEB

The SpatialExperiment objects must first be converted to Seurat objects.

```{r}
list.dirs.depth.n <- function(p, n) {
  res <- list.dirs(p, recursive = FALSE)
  if (n > 1) {
    add <- list.dirs.depth.n(res, n-1)
    c(res, add)
  } else {
    res
  }
}
files <- list.dirs.depth.n("/stornext/Projects/score/Analyses/G000218_spatial_benchmarking_study/extdata/SpaceRanger", 2)
run <- data.frame("709"="V42L05-390", "713"="V42L05-390")

seu <- function(spe) {
  names <- make.unique(rownames(spe))
  rownames(spe) <- names
  
  # get images
  img <- Read10X_Image(paste0(files[str_detect(files,paste0(run[,paste0("X", unique(spe$sample_id))],".*",unique(spe$sample_id)))], "/outs/spatial"), image.name = "tissue_hires_image.png")
  
  s <- CreateSeuratObject(
    counts = counts(spe),
    assay = "RNA",
    meta.data = as.data.frame(colData(spe)))
    s@meta.data <- s@meta.data[,c(1:3,5:6)]
    colnames(s@meta.data)[4:5] <- c("row","col")
    
  return(s)
}
seu_list <- list(seu(spe_709), seu(spe_713))
```

"First, we perform principal component analysis (PCA) on the combined log-transformed expression matrix to obtain the top principal components (PCs). Second, by taking the top PCs as input, we perform the integrative spatial clustering analysis using iSC-MEB that produces the soft cluster assignment as output. Finally, we can perform downstream analysis using the clustering result from the second step."

```{r}
library(iSC.MEB)
nnsvg_gene_list <- str_replace(df_summary$gene_id,"_","-") # match Seurat gene name format
obj <- CreateiSCMEBObject(seu_list, customGenelist = nnsvg_gene_list) # using custom gene list from nnSVG

# Add adjacency matrix list for a iSCMEBObj object to prepare for iSC.MEB model fitting.
obj <- CreateNeighbors(obj, platform="Visium")

# run PCA to get low dimensional embeddings
obj <- iSC.MEB::runPCA(obj, npcs = 10, pca.method = "APCA")

# Add a model setting in advance for an iSCMEBObj object.
obj <- SetModelParameters(obj, verbose = TRUE, maxIter=20, coreNum=4)
```

For function iSCMEB, users can specify the number of clusters K or set K to be an integer vector by using modified MBIC(BIC) to determine K. Here, we use user-specified number of clusters. 

```{r}
# 9 69347
# 10 68887
# 11 69274

# couldn't do K=m:n as in the tutorial - C stack usage over the limit - run iSCMEB() with each K and plot individually
obj <- iSCMEB(obj, K=8)
SelectKPlot(obj, criteria = "MBIC")
```

## Visualisation

```{r, fig.cap= "Spatial scatter plot."}
cols = c("#fb8072", "#bebada",  "#8dd3c7","#B5E8F9","#ffffb3",  "#b3de69", "#fdb462", "#FAABAB", "#fdd900", "#F9C8ED", "#55A5E8")
p1 <- SpaHeatMap(obj, item = "cluster", plot_type = "Scatter", nrow.legend = 1, no_axis = TRUE, cols = cols, point_size=1.75, combine=TRUE)
p1
```

Next, the user can visualize the inferred embeddings for biological effects between cell/domain types using two components from either tSNE or UMAP. Here, we demonstrate the clustering and batch remove performance by UMAP plot.

```{r, fig.cap="UMAP plot."}
obj <- CalculateUMAP(obj, reduction = "iSCMEB", n_comp = 2)

umap2 <- LowEmbedPlot(obj, item = "cluster", reduction = "UMAP2", point_size = 0.3)
umap3 <- LowEmbedPlot(obj, item = "batch", reduction = "UMAP2", point_size = 0.3)
umap2 + umap3
```

## DE

The IntegrateSpaData function can integrate multiple SRT data based on our results, function doDEG provide top differential expression genes, and function doHeatmap plot heatmap.

DE between spatial clusters.

```{r, fig.cap="Heatmap of combined DE analysis.",layout="l-body-outset",fig.width=10, fig.height=8}
seuInt <- IntegrateSpaData(obj,"mouse") # "mouse" indicates which species' housekeeping genes to use
top10 <- doDEG(seuInt, topn = 10)
p4 <- doHeatmap(seuInt, top10$gene)
p4
```

## Over-clustering

```{r}
spat_clust <- obj@resList@idents
for (i in 1:2) {
  spat_clust[[i]][spat_clust[[i]]==4] <- 2
  spat_clust[[i]][spat_clust[[i]]==5] <- 4
  spat_clust[[i]][spat_clust[[i]]==6] <- 5
  spat_clust[[i]][spat_clust[[i]]==7] <- 6
  spat_clust[[i]][spat_clust[[i]]==8] <- 7
}
obj@resList@idents <- spat_clust
```

## Visualisation

Next, the user can visualize the inferred embeddings for biological effects between cell/domain types using two components from either tSNE or UMAP. Here, we demonstrate the clustering and batch remove performance by UMAP plot.

```{r, fig.cap="UMAP plot."}
obj <- CalculateUMAP(obj, reduction = "iSCMEB", n_comp = 2)

p2 <- LowEmbedPlot(obj, item = "cluster", reduction = "UMAP2", point_size = 0.3)
p3 <- LowEmbedPlot(obj, item = "batch", reduction = "UMAP2", point_size = 0.3)
p2 + p3

# pdf(here("analysis","output","FFPE_CA_multi-sample_UMAP.pdf"), height=4, width=5)
p2
# dev.off()

# pdf(here("analysis","output","FFPE_CA_multi-sample_UMAP_batch.pdf"), height=4, width=5)
p3
# dev.off()

# saveRDS(obj,here("analysis","output","RDS","FFPE_CA_iSCMEB_obj.RDS"))
```

## DE

The IntegrateSpaData function can integrate multiple SRT data based on our results, function doDEG provide top differential expression genes, and function doHeatmap plot heatmap.

DE between spatial clusters.

```{r, fig.cap="Heatmap of combined DE analysis.", ,layout="l-body-outset",fig.width=10, fig.height=8}
seuInt <- IntegrateSpaData(obj,"mouse") # "mouse" indicates which species' housekeeping genes to use
top10 <- doDEG(seuInt, topn = 10)
p4 <- doHeatmap(seuInt, top10$gene)
p4
```


# Results and annotation

## Spatial marker gene expression

```{r, fig.cap="B marker gene expression.",layout="l-body-outset",fig.width=10, fig.height=10}
for (i in 1:length(b_markers)) {
  ix_gene <- which(rowData(spe)$symbol %in% b_markers[i])

  df <- as.data.frame(cbind(
  colData(spe), 
  spatialCoords(spe), 
  gene = counts(spe)[ix_gene, ]))

  plot <- ggplot(df, aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = gene)) + 
  facet_wrap(~ sample_id, nrow = 1, scales = "free") + 
  geom_point(size = 0.6) + 
  scale_color_gradient(low = "gray80", high = "red", trans = "sqrt", 
                       name = "counts", breaks = range(df$gene)) + 
  scale_y_reverse() + 
  ggtitle(paste0(rowData(spe)$symbol[ix_gene], " expression")) + 
  theme_bw() + 
  theme(aspect.ratio = 1, 
        panel.grid = element_blank(), 
        plot.title = element_text(face = "italic"), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank())
    assign(paste0("b",i), plot)
}
b1+b2+b3+b4+b5
```

```{r, fig.cap="T marker gene expression.",layout="l-body-outset",fig.width=10, fig.height=10}
for (i in 1:length(t_markers)) {
  ix_gene <- which(rowData(spe)$symbol %in% t_markers[i])

  df <- as.data.frame(cbind(
  colData(spe), 
  spatialCoords(spe), 
  gene = counts(spe)[ix_gene, ]))

  plot <- ggplot(df, aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = gene)) + 
  facet_wrap(~ sample_id, nrow = 1, scales = "free") + 
  geom_point(size = 0.6) + 
  scale_color_gradient(low = "gray80", high = "red", trans = "sqrt", 
                       name = "counts", breaks = range(df$gene)) + 
  scale_y_reverse() + 
  ggtitle(paste0(rowData(spe)$symbol[ix_gene], " expression")) + 
  theme_bw() + 
  theme(aspect.ratio = 1, 
        panel.grid = element_blank(), 
        plot.title = element_text(face = "italic"), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank())

  assign(paste0("t",i), plot)
}
t1+t2+t3+t4+t5
```


## Spatial clusters

Get spots in the same cell type in each sample

```{r}
deg <- doDEG(seuInt, topn=nrow(seuInt))
```

```{r,layout="l-body-outset",fig.width=10, fig.height=10}
colData(spe)$cluster <- unlist(spat_clust)

for (i in 1:8) {
  df <- as.data.frame(cbind(
    colData(spe),
    spatialCoords(spe)))
  
  df[df$cluster==i,"fill"] <- TRUE
  df[df$cluster!=i,"fill"] <- FALSE

  plot <- ggplot(df, aes(x = pxl_col_in_fullres, y = -pxl_row_in_fullres, color = fill)) + 
  facet_wrap(~ sample_id, nrow = 1, scales = "free") + 
  geom_point(size = 0.5) + 
  scale_colour_manual(values=c("TRUE"="red", "FALSE"="gray90")) +
  ggtitle(paste("Cluster",i)) + 
  theme_bw() + 
  theme(aspect.ratio = 1, 
        panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        strip.background = element_rect(fill = "#DCEFFE", color = "#DCEFFE"),
                          strip.text = element_text(color = "#074D6C", face = "bold", size = 15),
          panel.background = element_rect(fill = NA)) 

  assign(paste0("c",i), plot)
}
wrap_plots(c1,c2,c3,c4,c5,c6,c7,c8,nrow=3)


# png(here("analysis","output","FFPE_CA_cluster7.png"),height=2.5,width=7.5,units="in",bg="transparent", res=300)
# c7
# dev.off()
```

### Gene expression per spot

```{r}
# gene expr per spot
GeneExprSum_spot <- function(spatial_experiment_object, genes, colname) {
  common_genes <- intersect(rowData(spatial_experiment_object)$symbol, genes)
  spatial_experiment_object <- spatial_experiment_object[common_genes, ]
  genes_summed <- colSums(as.matrix(counts(spatial_experiment_object)))
  colData(spatial_experiment_object)[[colname]] <- genes_summed
  return(spatial_experiment_object)
}

for (i in 1:7) {
  df <- as.data.frame(cbind(
    colData(spe),
    spatialCoords(spe)))
  df$genes_sum <- GeneExprSum_spot(spe,t_markers,"genes_sum")$genes_sum
  
  df[df$cluster==i,"fill"] <- TRUE
  df[df$cluster!=i,"fill"] <- FALSE

plot <- ggplot(df, aes(x = pxl_col_in_fullres, y = -pxl_row_in_fullres,)) + 
  facet_wrap(~ sample_id, nrow = 1, scales = "free") + 
  geom_point(data = subset(df, cluster!=i), aes(color = genes_sum), size = 0.8, color = "gray95") + #show tissue but grey out
  geom_point(data = subset(df, cluster==i), aes(color = genes_sum), size = 0.8) +
  scale_color_distiller(palette="Oranges",direction= 1) +
  theme_bw() + 
  theme(aspect.ratio = 1, 
        panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        strip.background = element_rect(fill = "#DCEFFE", color = "#DCEFFE"),
                          strip.text = element_text(color = "#074D6C", face = "bold", size = 15))

  assign(paste0("c",i), plot)
}
wrap_plots(c1,c2,c3,c4,c5,c6,c7,nrow=3)


# pdf(here("analysis","output","FFPE_CA_Tcell_exp.pdf"),height=2.5,width=7.5)
c7
# dev.off()

```

### Assign labels

```{r,layout="l-body-outset",fig.width=10, fig.height=8}
df <- as.data.frame(cbind(
    colData(spe),
    spatialCoords(spe)))
  
df[df$cluster==1,"fill"] <- "Red pulp"
df[df$cluster==2,"fill"] <- "B cell"
df[df$cluster==3,"fill"] <- "Germinal centre"
df[df$cluster==4,"fill"] <- "Neutrophil"
df[df$cluster==5,"fill"] <- "Red pulp"
df[df$cluster==6,"fill"] <- "Plasma cell"
df[df$cluster==7,"fill"] <- "T cell"

# Define the order of clusters
cluster_order <- c("B cell", "Germinal centre", "Neutrophil", "Red pulp", "Plasma cell", "T cell")

# Convert 'fill' column to a factor and specify the levels
df$fill <- factor(df$fill, levels = cluster_order)

pal <- RColorBrewer::brewer.pal(6,"Set2")[c(1,3,5,4,6,2)]

labelled_plot <- ggplot(df, aes(x = pxl_col_in_fullres, y = -pxl_row_in_fullres, color = fill)) + 
  facet_wrap(~ sample_id, nrow = 1, scales = "free") + 
  geom_point(size = 0.8) + 
  scale_color_manual(values=pal) +
  theme_bw() + 
  theme(aspect.ratio = 1, 
        panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        legend.text = element_text(size=11),
        legend.title = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(color = "#074D6C", face = "bold", size = 15)) +
  guides(color = guide_legend(override.aes = list(size = 5)))

# pdf(here("analysis","output","FFPE_CA_labelled.pdf"), height=3,width=9)
labelled_plot
# dev.off()
```

### Cluster score heatmap

```{r}
# function for logfc matrix
logfc_mtx <- function(spe, cluster_label, marker_genes_df) {
  args.grid <- expand.grid(
    zone = as.character(unique(marker_genes_df$zone)),
    cluster = as.character(unique(cluster_label))
  ) |>
    tibble::as_tibble()

  parallel::mcmapply(function(zone, cluster) {
    rows <- marker_genes_df[marker_genes_df$zone == zone, "gene"]
    cols <- cluster_label == cluster
    cluster_score <- sum(SingleCellExperiment::counts(spe[rows, cols])) / sum(cols)
    other_score <- sum(SingleCellExperiment::counts(spe[rows, !cols])) / sum(!cols)
    tibble::tibble_row(
      zone = zone,
      cluster = cluster,
      logfc = log2(cluster_score / other_score)
    )
  }, args.grid$zone, args.grid$cluster, mc.cores = 4, SIMPLIFY = F) |>
    dplyr::bind_rows()
}

# markers df 
anno <- biomaRt::select(mart, keys=rownames(spe), keytype="external_gene_name",columns=c("external_gene_name","ensembl_gene_id"))
colnames(anno) <- c("symbol", "EnsembleID")

zone_markers_df <- 
list(
    "Macrophage" = c("Cd274", "Marco", "Csf1r", "Adgre1", "Cd209b", "Cd206", "Cd80", "Mac1", "Cd68")[c(1, 3, 4, 5, 7, 9)],
    "B cell" = c("Cd19", "Cd22", "Ighd", "Cd5"),
    "Germinal centre" = c("Cxcr4", "Cd83", "Bcl6", "Rgs13", "Aicda"),
    "Neutrophil" = c("S100a9", "S100a8", "Ngp"),
    "Erythrocyte" = c("Car2", "Car1", "Klf1"),
    "Plasma cell" = c("Cd38", "Cd138", "Xbp1", "Irf4", "Prdm1", "Cd27", "Cd319", "Mum1")[c(1, 3, 4, 5, 6)],
    "T cell" = c("Trac", "Cd3d", "Cd4", "Cd3e", "Cd8a"),
    "Red pulp" = c("Ifitm3", "C1qc", "Hmox1", "Hba-a1", "Klf1"),
    "Marginal zone" = c("Marco", "Lyz2", "Ighd", "Igfbp7", "Igfbp3", "Ly6d"),
    "White pulp" = c("Ighd", "Cd19", "Trac", "Trbc2")
  ) |>
    stack() |>
    setNames(c("gene", "zone")) |>
    merge(rowData(spe), # any spe with all EnsembleIDs and symbols for the markers should be fine
      by.x = "gene",
      by.y = "symbol"
    ) |>
    na.omit()

# plot
# spe_wtctrl
# cluster_wtctrl: factor same length as spe's cols
colData(spe)$cluster <- unlist(spat_clust)
cluster <- colData(spe)$cluster
p <- logfc_mtx(spe, cluster, zone_markers_df) |>
        ggplot(aes(
          x = factor(cluster, levels = 1:7),
          y = factor(zone, levels = rev(levels(zone_markers_df$zone))), fill = logfc
        )) +
        geom_tile(
          color = "white",
          lwd = 0.5,
          linetype = 1
        ) +
        theme_void() +
        theme(
          axis.title = element_blank(), plot.title = element_text(hjust = 0.5),
          axis.text = element_text(size = 12), legend.title = element_blank(),
          legend.text = element_text(size = 12)
        ) +
        geom_text(aes(label = format(round(logfc, digits = 1), nsmall = 1)), color = "black", size = 4) +
        scale_fill_distiller(palette = "RdBu") +
        coord_fixed()
      # ggsave(file.path("output", "WT_CTL_heatmap.jpg"), p, width = 15, height = 15, dpi = 600, units = "cm")
      # file.path("output", "WT_CTL_heatmap.jpg")

# pdf(here("analysis","output","FFPE_CA_cluster_score_heatmap.pdf"),height=5,width=6)
p
# dev.off()
```

## HVGs vs nnSVG

HVGs were selected using `scran` methods and purely defined on molecular features i.e. gene expression, not taking any spatial information into account.

```{r}
library(UpSetR)
hvg_709 <- hvg_709[order(hvg_709$p.value),]
hvg_713 <- hvg_713[order(hvg_713$p.value),]

list_input <- list(SVG_multi=rownames(df_summary), 
                   # SVG_709=names(res_ranks[,"709"][order(res_ranks[,"709"])]),
                   # SVG_713=names(res_ranks[,"713"][order(res_ranks[,"713"])]),
                   HVG_709=hvg_709$X, 
                   HVG_713=hvg_713$X)

# returns number of overlaps
top_n <- function(n) {
  top <- data.frame(list_input[[1]][1:n],
                    list_input[[2]][1:n],
                    list_input[[3]][1:n])#,
                    # list_input[[4]][1:n],
                    # list_input[[5]][1:n])
  length(Reduce(intersect, top))
}

n_overlap <- sapply(1:length(rownames(df_summary)), top_n)
max_index <- which.max(n_overlap / 1:length(rownames(df_summary))) # index of highest overlap - 541
n_overlap[max_index] / max_index # highest overlap - 0.76

# upset plot of the highest overlap
print(paste("Highest proportion of overlap:",round(n_overlap[which.max(n_overlap)],2)))
top <- list(SVG_multi=rownames(df_summary)[1:max_index],
            # SVG_709=names(res_ranks[,"709"][order(res_ranks[,"709"])])[1:which.max(n_overlap)],
            # SVG_713=names(res_ranks[,"713"][order(res_ranks[,"713"])])[1:which.max(n_overlap)],
            HVG_709=hvg_709$X[1:max_index],
            HVG_713=hvg_713$X[1:max_index])

# pdf(here("analysis","output","SVG_HVG_overlap.pdf"),height=5,width=7)
upset(fromList(top), 
      order.by = "freq",
      keep.order=TRUE,
      nsets=5,
      set_size.show=FALSE,
      text.scale=c(1.5, 1.5, 1.5, 1.5, 1.5, 2))
# dev.off()
```


# Deconvolution

```{r}
library(spacexr)
sce <- readRDS(here("analysis","output","RDS","sce.labelled.RDS"))
counts <- counts(sce)
rownames(counts) <- make.names(rownames(counts), unique=TRUE)

# process_cell_type_info error: need a minimum of 25 cells for each cell type in the reference
# fix: collapse into groups, myeloid: macrophage, monocyte, granulocyte: basophil, neutrophil, b cell: b cells, b cells, pro
cell_types <- as.character(sce$cluster_label); names(cell_types) <- sce$Barcode # create cell_types named list
# cell_types[cell_types %in% c("Macrophages","Monocytes","Stem cells","Granulocytes")] <- "Haematopoietic"
cell_types[cell_types %in% c("DC","NK cells","ILC")] <- "Immune"
cell_types <- as.factor(cell_types) # convert to factor data type
nUMI <- sce$sum; names(nUMI) <- sce$Barcode # create nUMI named list

### Create the Reference object
reference <- Reference(counts, cell_types, nUMI)
```

```{r, eval=FALSE}
for (i in 1){ # had to run separately
  hcc_st_s = seu_list[[i]]
  st_count = hcc_st_s@assays$RNA@counts
  st_coord = hcc_st_s@meta.data[,c('row','col')]
  nUMI <- colSums(st_count)
  puck <- SpatialRNA(st_coord, st_count, nUMI)
  output_name = paste0('FFPE_CA_RCTD_',i,"_grouped.RDS")
  myRCTD <- create.RCTD(puck, reference, max_cores = 2)
  myRCTD <- run.RCTD(myRCTD, doublet_mode = 'multi')
  assign(paste0("myRCTD_",i),myRCTD)
  saveRDS(myRCTD,here("analysis","output","RDS",output_name))}
```

```{r}
myRCTD_1 <- readRDS(here("analysis","output","RDS","FFPE_CA_RCTD_1_grouped.RDS"))
myRCTD_2 <- readRDS(here("analysis","output","RDS","FFPE_CA_RCTD_2_grouped.RDS"))

# get confident cell type proportions for each spot - sub_weights already normalised to sum to 1
for (i in 1:2) {
  all_weights <- list()
  myRCTD <- get(paste0("myRCTD_",i))
  weights <- data.frame(matrix(ncol=9))
  colnames(weights) <- levels(cell_types)
  for (j in 1:length(myRCTD@results)) {
    a <- data.frame(t(myRCTD@results[[j]]$sub_weights))
    colnames(a) <- myRCTD@results[[j]]$cell_type_list
    all_weights <- c(all_weights, list(a))
  } 
  weights <- dplyr::bind_rows(all_weights)
  weights[is.na(weights)] <- 0
  rownames(weights) <- rownames(get(paste0("myRCTD_",i))@spatialRNA@coords)
  assign(paste0("weights_",i), weights)
}

cell_type_names <- c("B cells","T cells","Erythrocytes","Myeloid")
for (j in 1:2) {
  assign(paste0("plot_",j), plot_weights("Erythrocytes", get(paste0("myRCTD_",j))@spatialRNA, get(paste0("weights_",j))))
}

# Create an empty list to store data frames
df_list <- list()

for (j in 1:2) {
  weights <- get(paste0("weights_",j))
  spatialRNA <- get(paste0("myRCTD_",j))@spatialRNA
  
  # Create a data frame for ggplot
  df <- data.frame(
    x = spatialRNA@coords[,1],
    y = spatialRNA@coords[,2],
    weight = weights[,"T cells"], # Erythrocytes for supplementary figure S7
    source = ifelse(j == 1, "709", "713")  # Add source column
  )
  
  # Add the data frame to the list
  df_list[[j]] <- df
}

# Combine the data frames
df_combined <- do.call(rbind, df_list)

deconv_plot <- ggplot(df_combined, aes(x = y, y = -x)) + 
  facet_wrap(~ source, nrow = 1, scales = "free") + 
  geom_point(data = subset(df_combined, weight < 0.2), aes(color = weight), size = 0.8, color = "gray90") + # add low weights to show tissue but grey out
  geom_point(data = subset(df_combined, weight >= 0.2), aes(color = weight), size = 0.8) +
  scale_color_gradientn(colors = pals::kovesi.rainbow(20), limits=c(0,1)) + # copy over spacexr's weights plotting
  theme_bw() + 
  theme(aspect.ratio = 1, 
        panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        strip.background = element_rect(fill = "#DCEFFE", color = "#DCEFFE"),strip.text = element_text(color = "#074D6C", face = "bold", size = 15))

# pdf(here("analysis","output","FFPE_CA_deconv.pdf"),height=2.5,width=7.5)
deconv_plot
# dev.off()
```


# Session info {.appendix}

The analysis and this document were prepared using the following software:

<details>

```{r}
devtools::session_info()
```

</details>
**Bioconductor version**: `r BiocManager::version()`




